God in Coq: \\ Modal Logic in Coq \\ and Scott's Version of G\"{o}del's Proof of God's Existence \\ {\small (\emph{Rough Diamond Proof Pearl})}

### 2014-01-31 Papers/ITP/GodProof-ITP.tex
Improve paper

A crucial aspect of modal logics \cite{ModalLogic} is that the so-called \emph{necessitation rule} allows $\Box A$ to be derived if $A$ is a theorem, but $A \imp \Box A$ is not necessarily a theorem. Naive attempts to define the modal operators $\Box$ and $\Diamond$ may easily be unsound in this respect. To avoid this issue, the \emph{possible world semantics} of modal logics can be explicitly embedded into higher-order logics \cite{J23,B9}. To this aim, a type for worlds must be declared and modal propositions should be not of type \texttt{Prop} but of a lifted type that depends on possible worlds. Possible worlds are connected by an accessibility relation.

The embedding is related to labeling techniques
\cite{Labels}. However, the expressiveness of higher-order logic is
exploited here to encode the labels within the logic language itself.

@@

\noindent
To maximally preserve user intuition in interactive modal logic theorem
proving, the embedding via the possible world semantics should be as
transparent as possible to the user. Fortunately, basic Coq tactics
such \texttt{intro}, \texttt{apply} and \texttt{split} automatically
unfold the modal definitions. They can thus be used with modal
connectives and quantifiers just as they are used with the usual
connectives and quantifiers.  For the modal operators, however, the
following simple tactics have been implemented, in order to allow the
user to work with the concepts of \emph{necessity} and
\emph{possibility} without having to unfold the definitions of modal
operators and think in terms of possible worlds.

@@

The successful formalization of Scott's version of G\"odel's proof of
God's existence indicates that the \emph{embedding} of modal logics
into higher-order logics via the \emph{possible world semantics} is a
viable approach for interactive theorem proving within modal logics.
The minimalistic implementation of the embedding (Section
\ref{sec:modal}) takes special care to hide the underlying possible
world machinery from the user.  An inspection of the proof scripts in
Section \ref{sec:proof} shows that this goal has been achieved. The
user does not have to explicitly bother about worlds and their mutual
reachability; the provided tactics for modalities do the job for him.
And for subgoals that do not involve modalities, the user has all the
usual interactive tactics at his/her disposal.
%The implementation of hints to allow automation tactics to fully cope with the embedding remains for future work. Nevertheless, at this stage a tactic such as \texttt{firstorder} is able to prove all lemmas and theorems in Section \ref{sec:modal} fully automatically.

@@

Related work includes formalizations of Anselm's argument by Rushby
\cite{Rushby}, and by Oppenheimer and Zalta \cite{Zalta}.  Obviously,
the application of theorem proving technology to the domain of
theoretical philosophy can --- as already pictured by Leibniz ---
be very fruitful for both areas. In fact, new results of
philosophical interest have been obtained by our
work~\cite{arXiv}. Moreover a first, basic infrastructure for
interactive and automated reasoning in higher-order modal logics has
been created and will be further improved in the future.

>>>>

A crucial aspect of modal logics \cite{ModalLogic} is that the so-called \emph{necessitation rule} allows $\Box A$ to be derived if $A$ is a theorem, but $A \imp \Box A$ is not necessarily a theorem. Naive attempts to define the modal operators $\Box$ and $\Diamond$ may easily be unsound in this respect. To avoid this issue, the \emph{possible world semantics} of modal logics can be explicitly embedded into higher-order logics \cite{J23,B9}. The embedding is related to labeling techniques
\cite{Labels}. However, the expressiveness of higher-order logic is
exploited here to encode the labels within the logic language itself. To this aim, a type for worlds must be declared and modal propositions should be not of type \texttt{Prop} but of a lifted type that depends on possible worlds. Possible worlds are connected by an accessibility relation.

@@

\noindent
To maximally preserve user intuition in interactive modal logic theorem
proving, the embedding via the possible world semantics should be as
transparent as possible to the user. Fortunately, basic Coq tactics
such as \texttt{intro}, \texttt{apply} and \texttt{split} automatically
unfold the modal definitions. They can thus be used with modal
connectives and quantifiers just as they are used with the usual
connectives and quantifiers.  For the modal operators, however, the
following simple tactics have been implemented, in order to allow the
user to work with the concepts of \emph{necessity} and
\emph{possibility} without having to unfold the definitions of modal
operators and think in terms of possible worlds.

@@

The successful formalization of Scott's version of G\"odel's proof of
God's existence indicates that the \emph{embedding} of modal logics
into higher-order logics via the \emph{possible world semantics} is a
viable approach for interactive theorem proving within modal logics.
The minimalistic implementation of the embedding (Section
\ref{sec:modal}) takes special care to hide the underlying possible
world machinery from the user.  An inspection of the proof scripts in
Section \ref{sec:proof} shows that this goal has been achieved. The
user does not have to explicitly bother about worlds and their mutual
reachability; the provided tactics for modalities do the job for him.
Moreover, for subgoals that do not involve modalities, the user has all the
usual interactive tactics at his/her disposal.
%The implementation of hints to allow automation tactics to fully cope with the embedding remains for future work. Nevertheless, at this stage a tactic such as \texttt{firstorder} is able to prove all lemmas and theorems in Section \ref{sec:modal} fully automatically.

@@

Related work includes formalizations of Anselm's argument by Rushby
\cite{Rushby}, and by Oppenheimer and Zalta \cite{Zalta}.  Clearly,
the application of theorem proving technology to the domain of
theoretical philosophy can --- as already pictured by Leibniz ---
be very fruitful for both areas. In fact, new results of
philosophical interest have been obtained by our
work~\cite{arXiv}. Moreover a first, basic infrastructure for
interactive and automated reasoning in higher-order modal logics has
been created and will be further improved in the future.
